---
output:
  pdf_document: default
  html_document:
    keep_md: yes
---

## Exponential Distribution Vs. the Central Limit Theorem

## Overview
This project explores the contrast between the observed and estimated means and variances of a large collection of randomly simulated exponential variables. This is achieved by comparing mean and variance values estimated by the Central Limit Theorem (CLT) to those observed through random simulation.

## Simulations

#### Load Required Libraries
```{r "libraries", message = F, warning = F}
library(grid)       # Plotting graphics (arrow)
library(ggplot2)    # Plotting (ggplot etc.)
```


#### Set Variables and Simulate Data
```{r "simulation"}
set.seed(123)
n <- 40
lambda <- 0.2
nSim <- 1000

exps <- NULL
means <- NULL
vars <- NULL

for (i in 1:nSim){
    exp <- rexp(n, lambda)
    exps <- cbind(exps, exp)
    means <- c(means, mean(exp))
    vars <- c(vars, sd(exp) ^ 2)
}

tMean <- 1 / lambda
tSD <- 1 / lambda
tSE <- tSD / sqrt(n)
tVar <- tSD ^ 2
```

The data were produced via 1,000 simulations (*nSim*) each generating 40 observations (*n*) of random exponentials using a rate of 0.2 (*lambda*). These values are applied via iteration of a loop to populate values of the expnonentials (*exps*), means (*means*) and variabilities (*vars*). 

Per the CLT both the estimated mean (*tMean*) and the estimated standard deviaion (*tSD*) are $(1/\lambda) = (1/0.2) = 5.0$.  Theoreteical standard error of the mean (*tSE*) is the standard deviation divided by the square root of the sample size = $\sigma / \sqrt{n} = 5.0 / \sqrt{1000} = .158$. Theoretical variance (*tVar*) is the square of the theoretical standard deviation $(5.0^2 = 25)$.  

## Sample Mean Vs. Theoretical Mean
```{r "sampleVtheoryMean", echo = F}
plot1 <- ggplot(
    as.data.frame(means)
    ,aes(x = means)
) + geom_histogram(
    alpha = .20
    ,binwidth = (range(means)[2] - range(means)[1]) / 20
    ,color = "black"
    ,aes(y = ..density..)    
) + geom_vline( # Sample distribution mean line
    xintercept = mean(means)
) + geom_segment( # Sample distribution mean arrow
    aes(
        x = mean(means) + 1
        ,xend = mean(means)
        ,y = max(density(means)$y)
        ,yend = max(density(means)$y) 
    )
    ,arrow = arrow(length = unit(.1,"in"), ends = "last")
) + geom_text( # Sample distribution mean label
    aes(
        x = mean(means) + 1
        ,y = max(density(means)$y)
        ,label = paste(
            "Sample distribution\nwith mean of"
            ,sprintf("%.5f", mean(means))
        )
    )
    ,color = "black"
    ,size = 4
    ,hjust = 0    
) + stat_function(
    fun = dnorm
    ,arg = list(mean = tSD)
    ,color = "red"
) + geom_vline( # Theoretical distribution variability line
    xintercept = tSD
    ,color = "red"
) + geom_segment( # Theoretical distribution variability arrow
    aes(
        x = tSD + 1
        ,xend = tSD
        ,y = max(density(means)$y) * .85
        ,yend = max(density(means)$y) * .85 
    )
    ,arrow = arrow(length = unit(.1,"in"), ends = "last")
    ,color = "red"
) + geom_text( # Theoretical distribution mean label
    aes(
        x = tSD + 1
        ,y = max(density(means)$y) * .85
        ,label = paste(
            "Theoretical distribution\nwith mean of"
            ,sprintf("%.5f",tMean)
        )
    )
    ,color = "red"
    ,size = 4
    ,hjust = 0    
) + labs(
    title = "Histogram of Exponential Sample\nvs. Theoretical Means"
    ,x = "Sample Means"
    ,y = "Density"
)
print(plot1)
```

The base plot in this figure is comprised of a histogram of the sample means with overlayed data including the actual sample mean as well as the theoretical distribution and mean.  This figure demonstrates that the sample mean (5.01191) is already  closely approximated by the theoretical mean (5.0) with a sample size of 1,000 simulations as described above.   


## Sample Vaiance vs. Theoretical Variance
```{r "sampleVtheoryVary", echo = F}
plot2 <- ggplot(
    as.data.frame(vars)
    ,aes(x = vars)
) + geom_histogram(
    alpha = .20
    ,binwidth = (range(vars)[2] - range(vars)[1]) / 20
    ,color = "black"
    ,aes(y = ..density..)    
) + geom_vline( # Sample distribution mean line
    xintercept = mean(vars)
) + geom_segment( # Sample distribution mean arrow
    aes(
        x = mean(vars) + 10
        ,xend = mean(vars)
        ,y = max(density(vars)$y)
        ,yend = max(density(vars)$y) 
    )
    ,arrow = arrow(length = unit(.1,"in"), ends = "last")
) + geom_text( # Sample distribution variability label
    aes(
        x = mean(vars) + 10
        ,y = max(density(vars)$y)
        ,label = paste(
            "Sample distribution\nwith variability mean of"
            ,sprintf("%.5f", mean(vars))
        )
    )
    ,color = "black"
    ,size = 4
    ,hjust = 0    
#  ) + stat_function(
#      fun = dnorm
#      ,arg = list(mean = tVar)
     ,color = "red"
) + geom_vline( # Theoretical distribution variability line
    xintercept = tVar
    ,color = "red"
) + geom_segment( # Theoretical distribution variability arrow
    aes(
        x = tVar + 10
        ,xend = tVar
        ,y = max(density(vars)$y) * .85
        ,yend = max(density(vars)$y) * .85
    )
    ,arrow = arrow(length = unit(.1,"in"), ends = "last")
    ,color = "red"
) + geom_text( # Theoretical distribution variability label
    aes(
        x = tVar + 10
        ,y = max(density(vars)$y) * .85
        ,label = paste(
            "Theoretical distribution\nwith variability mean of"
            ,sprintf("%.5f",tVar)
        )
    )
    ,color = "red"
    ,size = 4
    ,hjust = 0    
) + labs(
    title = "Density of Exponential Sample\nvs. Theoretical Variability"
    ,x = "Sample Variability"
    ,y = "Density"
)
print(plot2)
```

## Distribution
Show that the distribution is approximately normal.
Focus on the difference between the distribution of a large collection of random exponentials and the distribution of a large collection of averages of 40 exponentials. 

Via figures and text, explain how one can tell the distribution is approximately normal.

```{r "distribution", echo = F}
plot3 <- qplot(
    means
    ,geom = "density"
    ,main = "Density of Means for Exponential Variables:\nSample vs. Theoretical Values"
    ,xlab = "Sample Mean"
    ,ylab = "Density"
) + geom_vline( # Sample distribution mean line
    xintercept = mean(means)
) + geom_segment( # Sample distribution mean arrow
    aes(
        x = mean(means) + 1
        ,xend = mean(means)
        ,y = max(density(means)$y)
        ,yend = max(density(means)$y) 
    )
    ,arrow = arrow(length = unit(.1,"in"), ends = "last")
) + geom_text( # Sample distribution mean label
    aes(
        x = mean(means) + 1
        ,y = max(density(means)$y)
        ,label = paste(
            "Sample distribution\nwith mean of"
            ,sprintf("%.5f", mean(means))
        )
    )
    ,color = "black"
    ,size = 4
    ,hjust = 0
) + stat_function( # Theoretical (normal) distribution
    fun = dnorm
    ,color = "red"
    ,arg = list(mean = tMean)
) + geom_vline( # Theoretical distribution mean line
    xintercept = tMean
    ,color = "red"
) + geom_segment( # Theoretical distribution mean arrow
    aes(
        x = tMean + 1
        ,xend = tMean
        ,y = .4
        ,yend = .4 
    )
    ,arrow = arrow(length = unit(.1,"in"), ends = "last")
    ,color = "red"
) + geom_text( # Theoretical distribution mean label
    aes(
        x = tMean + 1
        ,y = .4
        ,label = paste(
            "Theoretical distribution\nwith mean of"
            ,sprintf("%.5f",tMean)
        )
    )
    ,color = "red"
    ,size = 4
    ,hjust = 0
)
print(plot3)
```


The CLT states that averages are approximately normal, with distributions
centered at the population mean and 
with standard deviation equal to the standard error of the mean

Standard Error of the mean = 
SEm = sigma / sqrt(N)