---
output:
  pdf_document: default
  html_document:
    keep_md: yes
---

## Title Analysis of the ToothGrowth Data from R Datasets Package

### Overview

This project explores the *ToothGrowth* data from the R *datasets* library by
performing some preliminary analysis on the data and drawing some initial
conclusions.


### Exploratory Data Analysis

```{r "setup", message = F, warning = F}
library(dplyr)      # Data manipulation (filter, mutate, group_by, etc.)
library(ggplot2)    # Plotting (qplot, ggplot etc.)
library(knitr)      # Dynamic Report Creation
library(datasets)   # R included sample data sets
data(ToothGrowth)
```

To setup the analysis we load the required libraries and the *ToothGrowth* data
set.  Per the R documentation for the *datasets* package [1], *ToothGrowth* is
described as follows:

> The response is the length of odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid).

```{r "exploratoryAnalysis"}
str(ToothGrowth)
summary(ToothGrowth)
table(ToothGrowth$supp)
table(ToothGrowth$dose)
table(ToothGrowth %>% select(supp, dose))

## Transformation/Renaming for clarity

names(ToothGrowth) = c("Length", "Suppliment", "Dose")
ToothGrowth$Dose = factor(ToothGrowth$Dose)
ToothGrowth$Suppliment <- factor(
    ToothGrowth$Suppliment
    ,labels = c("Orange Juice", "Ascorbic Acid")
)


##   summarize(n=n(),mn=mean(length),sd=sd(length)) %.%
##  mutate(se=sd/sqrt(n),LCI=mn+qnorm(0.025)*se,UCI=mn+qnorm(0.975)*se)

tgBySuppDose <- ToothGrowth %>%
    group_by(Suppliment, Dose) %>%
    summarize(Length = mean(Length), N = n(), SD = sd(Length)) %>%
    mutate(SE = SD/sqrt(N), CI = Length + qnorm(0.025) * SE) %>%
    arrange(Suppliment, Dose, desc(Length))
```

The exploratory analysis of the *ToothGrowth* data frame demonstrates that the
data consist of 30 observations each of *supp* OJ (Orange Juice) and VC 
(Ascorbic Acid); the 30 observations of each *supp* consist of 10 observations
each of the three *dose* levels (0.5, 1 and 2). We then create a data frame
grouped by Suppliment and Dose summarized by the means.

### Data Summary
```{r "summary"}
kable(tgBySuppDose
      ,format = "markdown"
      ,caption = "ToothGrowth by Suppliment and Dose"
)
```

Based on the summarized data we find that in general increasing the dosage of
Vitamin C leads to an increase in the mean of Tooth Length via both delivery
methods (orange juice and ascorbic acid).  While the highest supplied dosage
(2.0 mg) of each produce similar means (26.06 and 26.14, respectively) further
analysis for variability and confidence intervals is required.

### Confidence Intervals and Hypothesis Test

```{r "test"}
plot1 <- ggplot(
    ToothGrowth
    ,aes(
        x = Dose
        ,y = Length
        ,fill = Dose
    )
) + geom_boxplot() +
    facet_grid(.~Suppliment) +
    labs(
        title = "Analysis of ToothGrowth Data by Suppliment and Dose"
        ,x = "Dose (mg)"
        ,y = "Tooth Length"
    )
print(plot1)

plot2 <- ggplot(
    ToothGrowth
    ,aes(
        x = Dose
        ,y = Length
        ,color = Suppliment
        ,group = Suppliment
    )
) + geom_errorbar(
    aes(
        ymin = Length + se
        ,ymax = Length + se
    )
    ,color = "black"
    ,width=.1
    #,position = pd
)

print(plot2)
```

Outliers OJ 2 and Ascorbic Acid 1

### Conclusions

### Reference

[1] https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/ToothGrowth.html
