---
output: 
  html_document:
    keep_md: true
---

## Exponential Distribution Vs. the Central Limit Theorem

## Overview
This project explores the contrast between the observed and estimated means and variances of a large collection of randomly simulated exponential variables. This is achieved by comparing mean and variance values estimated by the Central Limit Theorem (CLT) to those observed through random simulation.

## Simulations

#### Load Required Libraries
```{r "libraries", message = F, warning = F}}
library(grid)       # Plotting graphics (arrow)
library(ggplot2)    # Plotting (qplot)
```


#### Set Variables and Simulate Data
```{r}
set.seed(123)
n <- 40
lambda <- 0.2
nSim <- 1000

exps <- NULL
means <- NULL
vars <- NULL

for (i in 1:nSim){
    exp <- rexp(n, lambda)
    exps <- cbind(exps, exp)
    means <- c(means, mean(exp))
    vars <- c(vars, sd(exp) ^ 2)
}

tMean <- 1 / lambda
tSD <- 1 / lambda
tVar <- tSD ^ 2
```

The data were produced via 1,000 simulations (*nSim*) each generating 40 observations (*n*) of random exponentials using a rate of 0.2 (*lambda*). These values are applied via iteration of a loop to populate values of the expnonentials (*exps*) and means (*means*). Per the CLT both the estimated mean (*tMean*) and the estimated standard deviaion (*tSD*) are (1/lambda) = (1/0.2) = 5.0.  Theoretical variance (*tVar*) is simply the square of the theoretical standard deviation (5.0^2 = 25).  

## Sample Mean Vs. Theoretical Mean

1. Show the sample mean and compare it to the theoretical mean of the distribution.
Sample Mean versus Theoretical Mean: Include figures with titles. In the figures, highlight the means you are comparing. Include text that explains the figures and what is shown on them, and provides appropriate numbers.

```{r "sampleVtheoryMean"}

plot1 <- ggplot(
    as.data.frame(means)
    ,aes(x = means)
) + geom_histogram(
    alpha = .20
    ,binwidth = (range(means)[2] - range(means)[1]) / 20
    ,color = "black"
    ,aes(y = ..density..)    
) + geom_vline( # Sample distribution mean line
    xintercept = mean(means)
) + geom_segment( # Sample distribution mean arrow
    aes(
        x = mean(means) + 1
        ,xend = mean(means)
        ,y = max(density(means)$y)
        ,yend = max(density(means)$y) 
    )
    ,arrow = arrow(length = unit(.1,"in"), ends = "last")
) + geom_text( # Sample distribution mean label
    aes(
        x = mean(means) + 1
        ,y = max(density(means)$y)
        ,label = paste(
            "Sample distribution\nwith mean of"
            ,sprintf("%.5f", mean(means))
        )
    )
    ,color = "black"
    ,size = 4
    ,hjust = 0    
) + stat_function(
    fun = dnorm
    ,arg = list(mean = tMean)
    ,color = "red"
) + geom_vline( # Theoretical distribution mean line
    xintercept = tMean
    ,color = "red"
) + geom_segment( # Theoretical distribution mean arrow
    aes(
        x = mean(means) + 1
        ,xend = tMean
        ,y = .4
        ,yend = .4 
    )
    ,arrow = arrow(length = unit(.1,"in"), ends = "last")
    ,color = "red"
) + geom_text( # Theoretical distribution mean label
    aes(
        x = mean(means) + 1
        ,y = .4
        ,label = paste(
            "Theoretical distribution\nwith mean of"
            ,sprintf("%.5f",tMean)
        )
    )
    ,color = "red"
    ,size = 4
    ,hjust = 0    
) + labs(
    title = "Density of Exponential Sample\nvs. Theoretical Means"
    ,x = "Sample Mean"
    ,y = "Density"
)
print(plot1)
```

## Sample Vaiance s. Theoretical Variance
Show how variable the sample is (via variance) and compare it to the theoretical variance of the distribution.

Include figures (output from R) with titles. Highlight the variances you are comparing. Include text that explains your understanding of the differences of the variances.
```{r "sampleVtheoryVary"}

```

## Distribution
Show that the distribution is approximately normal.
Focus on the difference between the distribution of a large collection of random exponentials and the distribution of a large collection of averages of 40 exponentials. 

Via figures and text, explain how one can tell the distribution is approximately normal.

```{r "distribution"}
plot3 <- qplot(
    means
    ,geom = "density"
    ,main = "Density of Means for Exponential Variables:\nSample vs. Theoretical Values"
    ,xlab = "Sample Mean"
    ,ylab = "Density"
) + geom_vline( # Sample distribution mean line
    xintercept = mean(means)
) + geom_segment( # Sample distribution mean arrow
    aes(
        x = mean(means) + 1
        ,xend = mean(means)
        ,y = max(density(means)$y)
        ,yend = max(density(means)$y) 
    )
    ,arrow = arrow(length = unit(.1,"in"), ends = "last")
) + geom_text( # Sample distribution mean label
    aes(
        x = mean(means) + 1
        ,y = max(density(means)$y)
        ,label = paste(
            "Sample distribution\nwith mean of"
            ,sprintf("%.5f", mean(means))
        )
    )
    ,color = "black"
    ,size = 4
    ,hjust = 0
) + stat_function( # Theoretical (normal) distribution
    fun = dnorm
    ,color = "red"
    ,arg = list(mean = tMean)
) + geom_vline( # Theoretical distribution mean line
    xintercept = tMean
    ,color = "red"
) + geom_segment( # Theoretical distribution mean arrow
    aes(
        x = mean(means) + 1
        ,xend = tMean
        ,y = .4
        ,yend = .4 
    )
    ,arrow = arrow(length = unit(.1,"in"), ends = "last")
    ,color = "red"
) + geom_text( # Theoretical distribution mean label
    aes(
        x = mean(means) + 1
        ,y = .4
        ,label = paste(
            "Theoretical distribution\nwith mean of"
            ,sprintf("%.5f",tMean)
        )
    )
    ,color = "red"
    ,size = 4
    ,hjust = 0
)
print(plot3)
```


The CLT states that averages are approximately normal, with distributions
centered at the population mean and 
with standard deviation equal to the standard error of the mean